---
title: "Machine Learning 1"
author: "Haroon Riyaz (PID A15377799)"
date: "2/14/2022"
output: pdf_document
---
# First: Kmeans()

Demo for using kmeans function in Base R! First make data with known structure!

```{r}
tmp <- c(rnorm(30,-3), rnorm(30,3))
x <- cbind(x=tmp ,y=rev(tmp))
plot(x)
```

Now we have made up data in 'x' let's see how kmeans works with this!

```{r}
k <- kmeans(x, centers = 2, nstart = 20)
k
```

> Q. How many points are in each cluster?

```{r}
k$size
```

> Q. How do we get to the cluster assignment?

```{r}
k$cluster
```

> Q. What are the cluster centers?

```{r}
k$centers
```

Now let's use results to plot data with the kmeans result!

```{r}
plot(x, col = k$cluster)
points(k$centers, col ="blue", pch = 15)
```

## hclust() or Hierarchial Clustering

We will cluster the same data 'x' with hclust! In this case, hclust requires a distance matrix as input.

```{r}
hc <- hclust(dist(x))
hc
```

Let's plot hclust result!

```{r}
plot(hc)
```

To get cluster membership vector we must "cut" tree with 'cutree()'.

```{r}
grps <- cutree(hc, h = 8)
grps
```

Now we plot our data with hclust results!

```{r}
plot(x, col=grps)
```






# Prinicipal Component Analysis 

## PCA of UK food data

Read data from website and try a few visualizations!

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url, row.names=1)
x
```
> Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

```{r}
dim(x)
```

> Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?

I prefer the alternative, second option as x <- x[,-1] may accidentally remove additional columns as the code block is repeated multiple times.

```{r}
x <- read.csv(url, row.names=1)
head(x)
```


```{r}
cols <- rainbow(nrow(x))
barplot(as.matrix(x), col = cols)
```

> Q3: Changing what optional argument in the *BELOW* (REVERSED ORDER) barplot() function results in the following plot (*ABOVE*)?

Setting beside=FALSE changes the graph to the one above since the bars are no longer beside one another. By default, bars are on top of each other unless beside is set to true.


```{r}
barplot(as.matrix(x), col = cols, beside = TRUE)
```

```{r}
pairs(x, col=cols)
```
> Q5: Generating all pairwise plots may help somewhat. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

Each graph represents one country being compared with the other in pairs (hence the name of the function pairs()). A given point that lies diagonal on a plot means that two values are similar to each other (example: countries drink similar amount of milk). Points that are not diagonal represent values that are not similar to each other (one country eats a lot of a specific food that another country doesn't).

> Q6. What is the main differences between N. Ireland and the other countries of the UK in terms of this data-set?

N. Ireland has more data points that are off the diagonal indicating that their eating habits differ the most compared to the other countries.

PCA to the rescue!
The main base R PCA function is 'prcomp()' and it will be trasnposed! 

```{r}
pca <- prcomp(t(x))
pca
```

Summary of what PCA is doing!

```{r}
summary(pca)
```

```{r}
attributes(pca)
```

To make new informative PCA plot (PCA scores) we access 'pca$x'

> Q7. Complete the code below to generate a plot of PC1 vs PC2. The second line adds text labels over the data points.

```{r}
plot(pca$x[,1], pca$x[,2])
text(pca$x[,1], pca$x[,2], colnames(x))
```

Color the plot!

>Q8. Customize your plot so that the colors of the country names match the colors in our UK and Ireland map and table at start of this document.

```{r}
country_cols <- c("orange", "red", "blue", "green")
plot(pca$x[,1], pca$x[,2], xlab = "PC1", ylab="PC2")
text(pca$x[,1], pca$x[,2], colnames(x),
     col=country_cols)
```
> Q9: Generate a similar ‘loadings plot’ for PC2. What two food groups feature prominantely and what does PC2 maninly tell us about?

```{r}
par(mar=c(10, 3, 0.35, 0))
barplot( pca$rotation[,2], las=2 )
```
Soft drinks and fresh potatoes are the two most prominent food groups! PC2 tells us about the variation between the data and shows strong patterns. 


## PCA of RNA-Seq Data

Read data from website

```{r}
url2 <- "https://tinyurl.com/expression-CSV"
rna.data <- read.csv(url2, row.names=1)
head(rna.data)
```
> Q10: How many genes and samples are in this data set?

```{r}
dim(rna.data)
```
There are 10 samples and 6 genes. 


```{r}
pca <- prcomp(t(rna.data))
summary(pca)
```

PCR plot of RNA-Seq Data

```{r}
plot(pca$x[,1], pca$x[,2], xlab = "PC1", ylab = "PC2")
```

```{r}
plot(pca$x[,1], pca$x[,2], xlab = "PC1", ylab = "PC2")
text(pca$x[,1], pca$x[,2], colnames(rna.data))
```


































