---
title: "Week 5 R functions"
author: "Haroon Riyaz (PID A15377799)"
date: "2/11/2022"
output: pdf_document
---

**Introduction to R functions**

Questions to answer:

> Q1. Write a function grade() to determine an overall grade from a vector of student homework
assignment scores dropping the lowest single score. If a student misses a homework (i.e. has an
NA value) this can be used as a score to be potentially dropped. Your final function should be
adquately explained with code comments and be able to work on an example class gradebook
such as this one in CSV format: “https://tinyurl.com/gradeinput” [3pts

```{r}
plot(1:10)
```



```{r}
# Example input vectors to start with
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
```


Follow the guidelines from class

-Write a working snipet of code that solves a simple problem

```{r}
# A straight forward mean (without dropping lowest score)
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)

mean(student1)
```
But ... we must drop lowest score. We must find lowest score first!

```{r}
# Which vector element is lowest
which.min(student1)
```
We want to exclude lowest value/score from main calculation.
```{r}
#Returns everything except 8th vector element
student1[-8]
```

Now we use which.min to return all elemnts except lowest value!

```{r}
# First working snippet 
mean(student1[-which.min(student1)])
```

What about other students?

We could use na.rm = TRUE argument but this will remove all NA values! Bad idea!

```{r}
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
#student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
mean(student2, na.rm = TRUE)
```

Another approach is to replace NA values with zero!


First we need to find NA elements! How do we do that?
```{r}
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
x <- student2

is.na(x)
which( is.na(x))
```

Now we have found the NA elements and mask them (replace with 0).

```{r}
# Not quite...
x[-which(is.na(x))] 
```

We must make the values zero!

```{r}
x[is.na(x)] <- 0
x
mean(x)
```

Recall we drop the lowest score!

```{r}
x[is.na(x)] <- 0
mean( x[-which.min(x)] )
```

Now we are almost done!
```{r}
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
x <- student3
x[is.na(x)] <- 0
mean( x[-which.min(x)] )
```

## Now we make our function

Take snippet and turn into function. Function has:

1) Function name "grade"
2) Input Arguments (student score vector)
3) The body (working snippet)

Using RStudio, code is selected 'Code -> Extract Function'
```{r}
#' Calculate average score for a vector of student scores that drops lowest score. Missing values are zero.
#'
#' @param x Numeric vector of HW scores
#'
#' @return Average score
#' @export 
#'
#' @examples
#' student <- c(90, 75, NA, 100)
#' grade(student)
#' 

grade <- function(x) {
  # Treat missing values as zero
  x[is.na(x)] <- 0
  # Excludes lowest score from the mean
  mean( x[-which.min(x)] )
}
```

```{r}
grade(student1)
grade(student2)
grade(student3)
```

This is great but we need to add comments now (explain to our future selves)!

Now we can use the function on the "real" whole class!
Use this CSV format: 
“https://tinyurl.com/gradeinput”

```{r}
url <- "https://tinyurl.com/gradeinput"
gradebook <- read.csv(url, row.names = 1)
# row.names = 1 puts students label in first gray column
```

```{r}
apply(gradebook, 1, grade)
# Applies grade function to student data set
```


> Q2. Using your grade() function and the supplied gradebook, Who is the top scoring student
overall in the gradebook? [3pts

To asnwer this we can use the apply function and save the results.

```{r}
results <- apply(gradebook, 1, grade)
sort(results, decreasing = TRUE)
```

```{r}
which.max(results)
```

> Q3. From your analysis of the gradebook, which homework was toughest on students (i.e. obtained the lowest scores overall? [2pts]

```{r}
gradebook
```

```{r}
mean.scores <- apply(gradebook,2, mean, na.rm=TRUE)
mean.scores
which.min(mean.scores)
```

```{r}
med.scores <- apply(gradebook,2, median, na.rm=TRUE)
med.scores
which.min(med.scores)
```

```{r}
boxplot(gradebook)
```

>Q4. Optional Extension: From your analysis of the gradebook, which homework was most
predictive of overall score (i.e. highest correlation with average grade score)? [1pt]

Are the final results (average score of each student) correlated with the results (scores) for individual HWs - the gradebook columns?

```{r}
masked.gradebook <- gradebook
masked.gradebook[is.na(masked.gradebook)] <- 0

masked.gradebook
```

Now look at correlation!

```{r}
cor(results, masked.gradebook$hw5)
```

```{r}
apply(masked.gradebook, 2, cor, x = results)
```

>Q5. Make sure you save your Rmarkdown document and can click the “Knit” button to generate
a PDF foramt report without errors. Finally, submit your PDF to gradescope. [1pt]

Knit the document to make a PDF




