---
title: "Project14"
author: "Haroon Riyaz (PID A15377799)"
date: "3/7/2022"
output: 
  pdf_document:
    latex_engine: xelatex
---

Read vaccination data

```{r}
vax <- read.csv("vacdata.csv")
head(vax)
```
> Q1. What column details the total number of people fully vaccinated? 

"persons_fully_vaccinated"

> Q2. What column details the Zip code tabulation area? 

"zip_code_tabulation_area"

> Q3. What is the earliest date in this dataset? 

2021-01-05

> Q4. What is the latest date in this dataset?

2022-03-01

~~~

Use the skim() function to get overview of data

```{r}
library(skimr)
numcol_skim <- skimr::skim(vax)
numcol_skim
```

> Q5. How many numeric columns are in this dataset? 

There are *9* numeric columns (excludes skim_variable and hist)

> Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column? 

```{r}
NA_PFV <- sum(is.na(vax$persons_fully_vaccinated))
NA_PFV
```

*18338* with current values (*18174* on lab sheet)

>Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)?

```{r}
# Divide # of missing values by # of rows, then round
round((NA_PFV/nrow(vax))*100, 2)
```

```{r}
# Lab Sheet Values
round((18174/105840)*100, 2)
```

*17.04%* with current values (*17.17%* on lab sheet)

> Q8. [Optional]: Why might this data be missing?

The data is missing since it represents people who haven't gotten the vaccine yet. 

~~~

## Using lubridate to Simplify Dates

```{r}
library(lubridate)
```

See Today's Date

```{r}
today()
```

```{r}
# Specify Date Format (ymd)

vax$as_of_date <- ymd(vax$as_of_date)
```

Now math can be done with the dates!

```{r}
today() - vax$as_of_date[1]
```

How many days the dataset spans

```{r}
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```

> Q9. How many days have passed since the last update of the dataset? 

```{r}
today() - vax$as_of_date[nrow(vax)]
```

6 days

> Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)?

```{r}
table(vax$as_of_date)
nrow(table(vax$as_of_date))
```
61 unique dates (lab sheet value different)

## Working With Zip Codes

```{r}
library("zipcodeR")
```

```{r}
geocode_zip('92037')
```

Calculate distance between 2 points with zip codes

```{r}
zip_distance('92037','92109')
```

Review Census Data

```{r}
reverse_zipcode(c('92037',"92109"))
```
## Focus on San Diego Area

```{r}
sd <- vax[vax$county == "San Diego", ]
```

Using dplyr

```{r}
library(dplyr)

sd <- filter(vax, county == "San Diego")

nrow(sd)
```

dplyr useful for subsetting

```{r}
sd.10 <- filter(vax, county == "San Diego" &
                  age5_plus_population > 10000)
```

> Q11. How many distinct zip codes are listed for San Diego County? 

```{r}
filtered_SD <- filter(vax, county == "San Diego")
length(unique(filtered_SD$zip_code_tabulation_area))
```
 *107* distinct zip codes in SD

> Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset? 

```{r}
maxSD12 <- which.max(filtered_SD$age12_plus_population)
filtered_SD$zip_code_tabulation_area[maxSD12]

```
*92154* is the zip code with largest 12+ pop.

> Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2022-02-22”? 

```{r}
PPFV_SD_02.22 <- filter(vax, county == "San Diego", as_of_date == "2022-02-22")

mean(PPFV_SD_02.22$percent_of_population_fully_vaccinated, na.rm = TRUE)
```

*70.42%*

> Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2022-02-22”?

```{r}
library(ggplot2)

ggplot(PPFV_SD_02.22, aes(percent_of_population_fully_vaccinated)) +
geom_histogram(bins = 12)
```

### Focus on UCSD

```{r}
ucsd <- filter(sd, zip_code_tabulation_area == "92037")
ucsd[1, ]$age5_plus_population
```

> Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area:

```{r}
ggplot(ucsd) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group = 1) +
  ylim(c(0,1)) +
  labs(x = "Date" , y = "Percent Vaccinated")
```

### Comparing to similar sized areas

```{r}
# CA areas with population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2022-02-22")

# head(vax.36)
```

> Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-02-22”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
mean_vax.36 <- mean(vax.36$percent_of_population_fully_vaccinated, na.rm = TRUE)
mean_vax.36
```

*73.34%* is the mean percentage of population fully vaccinated for zip codes with pop. as large as 92037

```{r}
ggplot(ucsd) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group = 1) +
  ylim(c(0,1)) +
  labs(x = "Date" , y = "Percent Vaccinated") +
  geom_hline(yintercept = mean_vax.36, linetype = "dashed", color = "blue")
```

> Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-02-22”?

```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```

> Q18. Using ggplot generate a histogram of this data.

```{r}
ggplot(vax.36, aes(percent_of_population_fully_vaccinated)) +
geom_histogram() + xlim(0,1)
```

> Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?

```{r}
vax %>% filter(as_of_date == "2022-02-22") %>%  
  filter(zip_code_tabulation_area =="92109") %>%
  select(percent_of_population_fully_vaccinated)

vax %>% filter(as_of_date == "2022-02-22") %>%  
  filter(zip_code_tabulation_area =="92040") %>%
  select(percent_of_population_fully_vaccinated)
```

Both ZIP code areas (92109 and 92040) are below the average value (0.7230 and 0.5513 < .7334)

> Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144.

```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)


ggplot(vax.36.all) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color="blue") +
  ylim(0,1) +
  labs(x = "Date", y = "Percent Vaccinated",
       title = "Vaccination rate across California",
       subtitle = "Only areas with a population above 36k are shown.") +
  geom_hline(yintercept = mean_vax.36, linetype = "dashed")
```

> Q21. How do you feel about traveling for Spring Break and meeting for in-person class afterwards?

I would feel great but this is my last quarter at UCSD so I won't be returning to in-person classes afterwards. 










































